---
title: "DATA 607 - Project 2 - Codebase"
author: "Kailot C. Harris"
date: "2026/10/08"
format:
  html:
    toc: true
    code-fold: true
    code-overflow: wrap
editor: visual
theme:
  dark: darkly
  light: flatly
---

# Introduction and Overview

The primary objective of this process is to transform three independent, wide data sets into formats ready for downstream analysis. Wide data sets are retrieved as .CSV files. The three selected data sets include synthetic Business Cost and Revenue data, 2022 US Census data, and World Bank Life Expectancy data. Each data set, after import, will be tidied according to standard conventions. Certain aspects of each data set will be analyzed: from the World Bank Data, rates of life expectancy across regions could be compared, while the US Census data could be analyzed for population growth momentum, and cost and revenue temporal trends could be determined for the business data set.

Anticipated challenges include initial import of unwieldy .CSV files. For instance, some may have header text that must be ignored when importing. Additionally, the World Bank data is separated into three distinct .CSV files, so joining these sets together as data frames may be necessary.

# Approach

1.  Import each data set
2.  Tidy each data set
3.  Analyze each data set
4.  Visualize each data set
5.  Present findings and recommendations for each data set

# Codebase

```{r packages}
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
```

## Data Import

### Data Sources

Data set 1 (2022 US Census data) was sourced from a discussion board post by [Anastasiia Gmyrina](https://brightspace.cuny.edu/d2l/le/1344368/discussions/threads/5805589/View). Data set 2 (Business Costs Revenue data) was sourced from a discussion board post by [Jocelyn Slater](https://brightspace.cuny.edu/d2l/le/1344368/discussions/threads/5777654/View). Data set 3 (World Bank Life Expectancy data) was sourced from a discussion board post by [Ummay Rukiya](https://brightspace.cuny.edu/d2l/le/1344368/discussions/threads/5847178/View). Data set 3 requires multiple imports for multiple .CSV files, and certain files have header information that requires skipping those rows.

```{r data-import}
#| echo: true
#| warning: false
#| message: false

url_1 <- "https://raw.githubusercontent.com/kailot-harris/DATA-607-Project-2/refs/heads/main/data/census_top10_states_2022_untidy.csv"

url_2 <- "https://raw.githubusercontent.com/kailot-harris/DATA-607-Project-2/refs/heads/main/data/5A_BusinessCostsRevenue.csv"

url_3a <- "https://raw.githubusercontent.com/kailot-harris/DATA-607-Project-2/refs/heads/main/data/API_SP.DYN.LE00.IN_DS2_en_csv_v2_460061.csv"

url_3b <- "https://raw.githubusercontent.com/kailot-harris/DATA-607-Project-2/refs/heads/main/data/Metadata_Country_API_SP.DYN.LE00.IN_DS2_en_csv_v2_460061.csv"

url_3c <- "https://raw.githubusercontent.com/kailot-harris/DATA-607-Project-2/refs/heads/main/data/Metadata_Indicator_API_SP.DYN.LE00.IN_DS2_en_csv_v2_460061.csv"

csv_data_1 <- readr::read_csv(file = url_1, 
                            show_col_types = FALSE,
                            progress = FALSE)

csv_data_2 <- readr::read_csv(file = url_2, 
                              show_col_types = FALSE,
                              progress = FALSE)

csv_data_3a <- readr::read_csv(file = url_3a, 
                               show_col_types = FALSE,
                               skip = 4,
                               progress = FALSE)

csv_data_3b <- readr::read_csv(file = url_3b, 
                               show_col_types = FALSE,
                               skip = 0,
                               progress = FALSE)

csv_data_3c <- readr::read_csv(file = url_3c, 
                               show_col_types = FALSE,
                               skip = 0,
                               progress = FALSE)
```

## Data structure before tidying

### Column Names

There are 5 column names for data set 1, displayed below:

```{r column-names-1}
names(csv_data_1)
```

There are 134 column names for data set 2, displayed below:

```{r column-names-2}
names(csv_data_2)

```

There are 71 column names in the first .CSV file for data set 3. The first 6 are displayed below:

```{r column-names-3a}
head(names(csv_data_3a),6)
```

There are 6 column names in the second .CSV file for data set 3, displayed below:

```{r column-names-3b}
names(csv_data_3b)
```

There are 5 column names in the third .CSV file for data set 3, displayed below:

```{r column-names-3c}
names(csv_data_3c)
```

### Row Count

There are 10 rows for data set 1:

```{r row-num-1}
nrow(csv_data_1)
```

There are 5 rows for data set 2:

```{r row-num-2}
nrow(csv_data_2)
```

There are 265 rows in the first .CSV file of data set 3:

```{r row-num-3a}
nrow(csv_data_3a)
```

There are 264 rows in the second .CSV file of data set 3:

```{r row-num-3b}
nrow(csv_data_3b)
```

There is only 1 row in the third .CSV file for data set 3:

```{r row-num-3c}
nrow(csv_data_3c)
```

### Representative Sample

The first 9 rows for the 5 columns from data set 1 are displayed below:

```{r rep-samp-1}
knitr::kable(head(csv_data_1,9))
```

The first 5 rows of the first 5 columns of data set 2 are displayed below:

```{r rep-samp-2}
knitr::kable(csv_data_2[1:5, 1:5])
```

The first 9 rows of the first 4 columns of the first .CSV file from data set 3 are displayed below:

```{r rep-samp-3a}
knitr::kable(csv_data_3a[1:9, 1:4])
```

The first 6 rows of 5 of the 6 columns of the second .CSV file from data set 3 are displayed below. One of the columns was removed prior to sampling because it contained new-line characters (e.g. "\\n"), causing the knitr table to crash when displaying. This column removal was accomplished by selecting all columns but the problematic 'SpecialNotes' column.

```{r rep-samp-3b}
csv_data_3b |>
  select(-SpecialNotes) |>
  head() |>
  knitr::kable()

```

The only row of the first 2 columns of the third .CSV file from data set 3 are displayed below. The remaining columns contained new-line characters (e.g. "\\n"), causing the knitr table to crash when displaying.

```{r rep-samp-3c}
knitr::kable(csv_data_3c[1:1, 1:2])
  

```

# Data Transformation

### Commented, executable code with explanations for each decision

# Analysis

### Description of the statistical or summarization approach

# Conclusions
